---
output:
  word_document: default
  pdf_document: default
  html_document: default
---
# Quantium Virtual Internship - Retail Strategy and Analytics - Task 1

*submitted by:* **Apoorv Rathaur**

## Solution for Data preparation and customer analytics task

*Setting the directory*
```{r}
setwd("E:/Datasets")
```

*Importing libraries*
```{r}
library(tidyverse)
library(data.table)
library(dplyr)
```

*Importing Datasets*
```{r}
tdata <- fread(paste0("QVI_transaction_data2.csv"))
cdata <- fread(paste0("QVI_purchase_behaviour.csv"))
```

## Exploratory Data Analysis

The first step in any analysis is to first understand the data. Let's take a look at each of the datasets provided.

```{r}
##Checking structure of the data
str(tdata)
```

We can see that the date column is in an integer format. Let's change this to a date format.

```{r}
#Changing date from numeric to Date format
tdata$DATE <- as.Date(tdata$DATE, origin = "1899-12-30")
```

We should check that we are looking at the right products by examining PROD_NAME.

```{r}
#Examining PROD_NAME column
summary.factor(tdata$PROD_NAME, maxsum = -10)

pw <- data.table(as.factor(unlist(strsplit(tdata[, tdata$PROD_NAME], " "))))
setnames(pw, 'words')
```

As we are only interested in words that will tell us if the product is chips or not, let's remove all words with
digits and special characters such as '&' from our set of product words. We can do this using grepl().

```{r}
#### Removing digits
pw <- pw[grepl("\\d", words) == FALSE, ]
#### Removing special characters
pw <- pw[grepl("[:alpha:]", words), ]
```

```{r}
#Sorting them in descending order
count1 <- pw %>% group_by(words) %>% summarise(n = n(), ) %>% arrange(desc(n))
count1
```

There are salsa products in the dataset but we are only interested in the chips category, so let's remove these

```{r}
#Removing SALSA products
tdata[, SALSA := grepl("salsa", tolower(PROD_NAME))]
tdata <- tdata[SALSA == FALSE, ][, SALSA := NULL]
```

```{r}
#Checking Null values
sum(is.na(tdata)) 
```


```{r}
##Handling Outliers
#omitting categorical variable to plot boxplot
bp <- tdata[,-c(1:6)]
boxplot(bp)
```
```{r}
bp[bp$PROD_QTY >= 100,]
```
```{r}
#checking the customer info
tdata[tdata$PROD_QTY >= 100,] 
```

There are two transactions where 200 packets of chips are bought in one transaction and both of these
transactions where by the same customer.

```{r}
#other transactions that customer made
tdata[tdata$LYLTY_CARD_NBR == 226000,] 
```
It looks like this customer has only had the two transactions over the year and is not an ordinary retail customer. The customer might be buying chips for commercial purposes instead. We'll remove this loyalty card number from further analysis.

```{r}
#Filtering out the customer based on the loyalty card number
tdata1 <- tdata[!tdata$LYLTY_CARD_NBR == 226000,]
```

That's better. Now, let's look at the number of transaction lines over time to see if there are any obvious data issues such as missing data.

```{r}
#Counting the number of transactions by date
tdata1 %>%
  group_by(DATE) %>%
  summarise(n = n())
```

Only 364 observations that mean there is a missing date.Let's create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a chart of number of transactions over time to find the missing date.

```{r}
#Create a sequence of dates and join this the count of transactions by date
dt <- data.frame(DATE = c(43282:43646))
dt$DATE <- as.Date(dt$DATE, origin = "1899-12-30")

#joining transactions data by count data of date
tdata1 <- right_join(tdata1, dt)
```

```{r}
#Missing Date
tdata1 %>%
  group_by(DATE) %>%
  summarise(n = n()) %>%
  filter(n <= 1)
```

Comes out that the missing date is "2018-12-25".

```{r}
#Summary of transaction count by date
dt <- tdata1 %>%
  group_by(DATE) %>%
  summarise(n = n()) %>%
  ungroup()
dt
```

```{r}
#Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

#Plotting transactions over time
ggplot(dt, aes(x = DATE, y = n)) +
  geom_line() +
  labs(x = "Months", y = "Number of transactions", title = "Transactions over time") +
  scale_x_date(breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see that there is an increase in purchases in December and a break in late December. Let's zoom in on this.

```{r}
#Filter to December and looking at individual days
dt %>%
  filter(between(DATE, as.Date("2018-12-01"), as.Date("2018-12-31"))) %>%
  ggplot(., aes(x = DATE, y = n)) +
  geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
  scale_x_date(breaks = "1 day") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see that the increase in sales occurs in the lead-up to Christmas and that there are zero sales on Christmas day itself. This is due to shops being closed on Christmas day.

Now that we are satisfied that the data no longer has outliers, we can move on to creating other features such as brand of chips or pack size from PROD_NAME. We will start with pack size.

```{r}
#We can work this out by taking the digits that are in PROD_NAME
tdata1[, PACK_SIZE := parse_number(PROD_NAME)]
```
```{r}
#checking if the pack sizes look sensible
dat <- tdata1[, .N, PACK_SIZE][order(PACK_SIZE)]
dat
```

The largest size is 380g and the smallest size is 70g - seems sensible!

```{r}
#plotting a histogram of PACK_SIZE
dat %>%
  ggplot(., aes(x =PACK_SIZE)) +
  xlab("Pack Size") + ylab("Frequency") +
    geom_histogram(binwidth = 50, col = "black")
```

Pack sizes created look reasonable and now to create brands, we can use the first word in PROD_NAME to
work out the brand name.

```{r}
#Creating a column which contains the brand of the product
library(tidyr)
tdata2 <- tdata1 %>%
  separate(PROD_NAME, into = c("BRAND", "Other")) #sep = "[^[:alnum:]]+" default
tdata2 <- tdata2[, -7]
```
```{r}
#### Clean brand names
tdata2[BRAND == "Dorito", BRAND := "Doritos"]
tdata2[BRAND == "Grain", BRAND := "GrnWves"]
tdata2[BRAND == "Infzns", BRAND := "Infuzions"]
tdata2[BRAND == "Natural", BRAND := "NCC"]
tdata2[BRAND == "Red", BRAND := "RRD"]
tdata2[BRAND == "Smith", BRAND := "Smiths"]
tdata2[BRAND == "Snbts", BRAND := "Sunbites"]
tdata2[BRAND == "Woolworths", BRAND := "WW"]
```
```{r}
#the results
dt2 <- tdata2 %>% 
  group_by(BRAND) %>%
  summarise(n = n())
```

## Examining Customer Data

```{r}
#basic summaries of the dataset, including distributions of any key columns
str(cdata)
```
```{r}
#Examining the values of lifestage and premium_customer
cdata %>% group_by(LIFESTAGE) %>% summarise(n = n())
```
```{r}
cdata %>% group_by(PREMIUM_CUSTOMER) %>% summarise(n = n())
```

As there do not seem to be any issues with the customer data, we can now go ahead and join the transaction and customer data sets together.

```{r}
#Merge transaction data to customer data
data <- merge(tdata2, cdata, all.x = TRUE)
data <- data[-1,] #removing the Christmas date
```

Let's also check if some customers were not matched on by checking for nulls.

```{r}
#Check for missing customer details
sum(is.na(data))
```
```{r}
data[is.null(LIFESTAGE), .N]
```
```{r}
data[is.null(PREMIUM_CUSTOMER), .N]
```

*Code to save dataset as a csv*
```{r}
write.csv(data, file = "QVI_data.csv", row.names = FALSE)
```

## Data analysis on customer segments

Calculating total sales by *LIFESTAGE* and *PREMIUM_CUSTOMER* and plotting the split by these segments to describe which customer segment contribute most to chip sales.

```{r}
# Total sales by LIFESTAGE and PREMIUM_CUSTOMER
Sales_data <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  mutate(SALES = sum(TOT_SALES)) %>%
  select(LIFESTAGE, PREMIUM_CUSTOMER, SALES) %>%
  distinct()%>%
  ungroup() %>%
  mutate(PercentofTotal = round((SALES/sum(SALES))*100, digits = 1))

#Plotting
Sales_data %>%
  ggplot(. , aes(x = LIFESTAGE, y = PercentofTotal, fill = PREMIUM_CUSTOMER)) +
  geom_bar(position = "dodge",stat = "identity", width = 0.8, color = "black", alpha = 0.6) +
    labs (x = "Lifestage", y = "Premium customer flag", title = "PROPORTION OF SALES") +
  scale_y_continuous(breaks = seq(0,10,2))+
  geom_text(aes(label = PercentofTotal), position=position_dodge(width = 0.8), vjust = -0.25, size = 2) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

*Let's see if the higher sales are due to there being more customers who buy chips.*
```{r}
#Number of customers by LIFESTAGE and PREMIUM_CUSTOMER
Customer_Count <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  mutate(CUSTOMERS = uniqueN(LYLTY_CARD_NBR)) %>%
  select(LIFESTAGE, PREMIUM_CUSTOMER, CUSTOMERS) %>%
  distinct() %>% ungroup() %>%
  mutate(PercentofTotal = round((CUSTOMERS/sum(CUSTOMERS))*100, digits = 1))

#Plotting
Customer_Count %>%
  ggplot(. , aes(x = LIFESTAGE, y = PercentofTotal, fill = PREMIUM_CUSTOMER)) +
  geom_bar(position = "dodge",stat = "identity", width = 0.6, color = "black", alpha = 0.6) +
  labs (x = "Lifestage", y = "Premium customer flag", title = "PROPORTION OF CUSTOMERS") +
  scale_y_continuous(breaks = seq(0,10,2))+
  geom_text(aes(label = PercentofTotal), position=position_dodge(width = 0.8), vjust = -0.25, size = 2) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

There are more Mainstream - young singles/couples and Mainstream - retirees who buy
chips. This contributes to there being more sales to these customer segments but
this is not a major driver for the Budget - Older families segment.

Higher sales may also be driven by more units of chips being bought per customer.
Let's have a look at this next.

```{r}
#Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER
Avg_Units <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  mutate(AVG = sum(PROD_QTY)/uniqueN(LYLTY_CARD_NBR)) %>%
  select(LIFESTAGE, PREMIUM_CUSTOMER, AVG) %>%
  distinct() %>% ungroup() %>%
  mutate(PercentofTotal = round((AVG/sum(AVG))*100, digits = 1))

#Plotting
Avg_Units %>%
  ggplot(. , aes(x = LIFESTAGE, y = PercentofTotal, weight = AVG, fill = PREMIUM_CUSTOMER)) +
  geom_bar(position = "dodge",stat = "identity", width = 0.6, color = "black", alpha = 0.5) +
  labs (x = "Lifestage", y = "Avg units per transaction", title = "UNITS PER CUSTOMER") +
  geom_text(aes(label = PercentofTotal), position=position_dodge(width = 0.8), vjust = -0.25, size = 2) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Older families and young families in general buy more chips per customer
Let's also investigate the average price per unit chips bought for each customer
segment as this is also a driver of total sales.

```{r}
#Average price per unit by LIFESTAGE and PREMIUM_CUSTOMER
Avg_Price <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  mutate(AVG = sum(TOT_SALES)/sum(PROD_QTY)) %>%
  select(LIFESTAGE, PREMIUM_CUSTOMER, AVG) %>%
  distinct() %>% ungroup() %>%
  mutate(PercentofTotal = round((AVG/sum(AVG))*100, digits = 1))

#Plotting
Avg_Price %>%
  ggplot(. , aes(x = LIFESTAGE, y = PercentofTotal, weight = AVG, fill = PREMIUM_CUSTOMER)) +
  geom_bar(position = "dodge",stat = "identity", width = 0.6, color = "black", alpha = 0.5) +
  labs (x = "Lifestage", y = "Avg price per unit", title = "PRICE PER UNIT") +
  geom_text(aes(label = PercentofTotal), position=position_dodge(width = 0.8), vjust = -0.25, size = 2) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Mainstream midage and young singles and couples are more willing to pay more per packet of chips compared to their budget and premium counterparts. This may be due to premium shoppers being more likely to buy healthy snacks and when they buy chips, this is mainly for entertainment purposes rather than their own consumption. This is also supported by there being fewer premium midage and young singles and couples buying chips compared to their mainstream counterparts.
As the difference in average price per unit isn't large, we can check if this difference is statistically different.

```{r}
#performing an independent t-test between mainstream vs premium and budget midage and
#young singles and couples
PricePerUnit <- data %>%
  mutate(price = TOT_SALES/PROD_QTY)
t.test(PricePerUnit[LIFESTAGE %in% c ("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES")
             & PREMIUM_CUSTOMER == "Mainstream", price],
        PricePerUnit[LIFESTAGE %in% c ("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES")
               & PREMIUM_CUSTOMER != "Mainstream", price],
        alternative = "greater")
```

The t-test results in a p-value of p-value < 2.2e-16, i.e. the unit price for mainstream, young and mid-age singles and couples are significantly higher than that of budget or premium, young and midage singles and couples.

## Deep dive into Mainstream, young singles/couples

```{r}
#Deep dive into Mainstream, young singles/couples
segment1 <- data[LIFESTAGE == "YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER == "Mainstream",]
other <- data[!(LIFESTAGE == "YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER == "Mainstream"),]

#Brand affinity compared to the rest of the population
quantity_segment1 <- sum(segment1$PROD_QTY)

quantity_other <- sum(other$PROD_QTY)

quantity_segment1_by_brand <- segment1 %>% group_by(BRAND) %>%
  summarise(targetSegment = sum(PROD_QTY)/quantity_segment1) %>% 
  select(BRAND, targetSegment)

quantity_other_by_brand <- other %>% group_by(BRAND) %>%
  summarise(other = sum(PROD_QTY)/quantity_other) %>%
  select(BRAND, other)

brand_proportions <- merge(quantity_segment1_by_brand, quantity_other_by_brand)
brand_proportions <- brand_proportions %>%
  mutate(affinityToBrand = targetSegment/other) %>% arrange(desc(affinityToBrand))
brand_proportions
```

We can see that :
. Mainstream young singles/couples are 22.8% more likely to purchase Tyrrells chips compared to the
  rest of the population
. Mainstream young singles/couples are 56% less likely to purchase Burger Rings compared to the rest
  of the population

Let's also find out if our target segment tends to buy larger packs of chips.

```{r}
#Preferred pack size compared to the rest of the population
quantity_segment1_by_PS <- segment1 %>% group_by(PACK_SIZE) %>%
  summarise(targetSegment = sum(PROD_QTY)/quantity_segment1) %>% 
  select(PACK_SIZE, targetSegment)

quantity_other_by_PS <- other %>% group_by(PACK_SIZE) %>%
  summarise(other = sum(PROD_QTY)/quantity_other) %>%
  select(PACK_SIZE, other)

Pack_proportions <- merge(quantity_segment1_by_PS, quantity_other_by_PS)
Pack_proportions <- Pack_proportions %>%
  mutate(affinityToPack = targetSegment/other) %>% arrange(desc(affinityToPack))
Pack_proportions
```

It looks like Mainstream young singles/couples are 26.8% more likely to purchase a 270g pack of chips compared to the rest of the population but let's dive into what brands sell this pack size.

```{r} data[PACK_SIZE == 270, unique(BRAND)]```

Twisties are the only brand offering 270g packs and so this may instead be reflecting a higher likelihood of purchasing Twisties.
